\ttitle

### 2015-08-18 Chapters/Introduction.tex
Huge commit following on from Donny conversation
- Change all references to before period
- Remove post 2011/2012 references from introduction - one 2012 reference left was published online in Aug 2011
- Rename title - 'investigation into clinical utility of CPET'
- Remove hyphenation of words
- Left justify with ragged right edge
- Remove bold formatting in some tables
- Minor grammar and style corrections in multiple places
- Change reference Chandrabalan 2013 to Chapter 2
- Create some new tables for body comp chapter based on Colin Richards paper - these dont work

Pancreatic ductal adenocarcinoma is the most common cancer of the pancreas. However, the head of the pancreas is anatomically related to several other epithelium lined structures that can also give rise to cancers. These include the distal common bile duct (cholangiocarcinoma), the duodenum (duodenal adenocarcinoma) and the ampulla (ampullary adenocarcinoma). Neuroendocrine tumours (NET) are a collection of a variety of neoplasia that originate from the endocrine portion of the pancreas. This milieu is complicated by other neoplasia such as intra-ductal papillary neoplasms (IPMN) as well as rare stromal tumours. Occasionally, chronic pancreatitis may present with features similar to pancreatic cancer and can be morphologically, radiologically and histologically difficult to differentiate from cancer.

Pancreatic cancer is the tenth most common cancer in the UK but the fifth most common cause of cancer death. Only 21\% survive beyond the first year and 5-year survival is 3\%. \parencite{cancerresearchuk_cancer_2014} The majority of patients (80-85\%) with pancreatic cancer present with inoperable disease.\parencite{cancerresearchuk_cancer_2014,sener_pancreatic_1999}

@@

Initial assessment includes a thorough history, clinical examination and blood tests including full blood count, renal function tests, liver function tests and tumour markers, especially CA 19-9. However, definitive diagnosis requires cross-sectional imaging in the form of a contrast-enhanced computerised tomogram (CECT) of the abdomen using a pancreas-specific protocol (a modified form of the portal-venous phase). CECT of the pancreas when combined with CT Thorax also provides accurate information on staging of the disease with regards to metastasis and this can be supplemented by further imaging such as Positron Emission Tomography (PET-CT) or contrast-enhanced Magnetic Resonance Imaging (MRI) of the liver in specific cases. CECT-pancreas is also useful for assessing local resectability with accurate information regarding vascular involvement. Endoscopic Ultrasound (EUS) is used in selected patients to assess vascular involvement and for obtaining tissue samples for histological examination. In jaundiced patients, endoscopic retrograde cholangio pancreatography (ERCP) plays an important role in the alleviation of jaundice by placing stents across the obstructed bile duct, accurate visualisation of the biliary anatomy as well as obtaining brushings from within the bile duct for cytological examination. The role of preoperative biliary drainage is discussed in more detail in Section \ref{sec:preoperative_biliary_drainage}.

\subsection{Treatment of pancreatic cancer}
Pancreaticoduodenectomy followed by adjuvant chemotherapy offers the only chance of cure in patients with resectable pancreatic cancer who are fit enough to undergo surgery. In patients with unresectable disease or who are not fit to undergo surgery, palliative chemotherapy plays a limited role in prolonging survival. Assessing the resectability of lesions of the head of the pancreas is discussed in the next section while the assessment of patient fitness and the impact of comorbidity are discussed in detail in section \ref{sec:comorbidity_risk_stratification} on p\pageref{sec:comorbidity_risk_stratification}.

\section{Surgical treatment of pancreatic cancer}
Pancreaticoduodenectomy remains a technically challenging and complex surgical procedure over a hundred years after its description. Excision of cancer of the head of the pancreas was first described as a two-stage operation by a German surgeon, Walther Kausch in 1909 at Augusta-Viktoria-Krankenhaus in Berlin-Schöneberg.\parencite{kausch_carcinom_1912}. The operation was further popularised initially as a two-stage procedure by Whipple\parencite{whipple_treatment_1935} before evolving into the current single stage operation by the 1950s.\parencite{whipple_rationale_1941,whipple_radical_1950}

\subsection{Resectability criteria}
\label{sec:resectability_criteria}
Resectable pancreatic cancer is defined as a tumour that does not involve the coeliac axis or the superior mesenteric artery and is not associated with distant metastatic disease. Tumours involving the portal vein or superior mesenteric vein are considered borderline resectable and can still be resected completely (R0) with en-bloc venous resection in selected cases.\parencite{fuhrman_rationale_1996} The role of neoadjuvant therapy such as chemotherapy with or without local radiotherapy and newer treatment modalities such as electroporation\parencite{bower_irreversible_2011}  and intra-peritoneal chemotherapy is the subject of ongoing clinical research in patients with locally advanced and/or metastatic disease. Early reports suggest that some of these patients can undergo potentially curative surgery after neo-adjuvant treatment.

@@

A roof-top incision was made to enter the abdominal cavity. After assessing the peritoneal cavity and liver for absence of metastatic disease, an early assessment was made for local resectability. This involved complete Kocherisation of the duodenum to assess the retroperitoneum. Both the superior mesenteric artery and coeliac axis were assessed early for tumour involvement (‘artery-first’ approach). The rest of the procedure was performed as described extensively elsewhere. The gastrocolic omentum was divided to enter the lesser sac. The superior mesenteric vein was identified and a retro-pancreatic tunnel was created between the pancreatic neck and the portal vein. If less than half the circumference of the superior mesenteric vein or portal vein was involved, an en-bloc resection was performed with vein repair at the same time. The hepatoduodenal ligament was dissected after a fundus-first cholecystectomy to isolate the common bile duct which was transected after ascertaining the hepatic artery anatomy. The gastro-duodenal artery was divided leaving a short segment on the common hepatic artery should the patient require embolisation of this vessel for post-pancreatectomy haemorrhage. The stomach (classical Whipple) or the first part of the duodenum (pylorus-preserving pancreaticoduodenectomy) were divided using a stapling device. The pancreatic neck was transected using Harmonic scalpel, diathermy or cold blade depending on surgeon preference. The jejunum was divided at the first jejunal arcade and the jejunal mesentery was divided using the Harmonic ultrasonic scalpel. The retroportal lamina was divided preserving the superior mesenteric artery.

Reconstruction was performed as follows: Either a pancreatico-jejunostomy was performed using 4-0 Biosyn sutures in a two-layer duct-to-mucosa technique or a pancreatico-gastrostomy was performed using 3/0 Biosyn sutures placed in a similar manner. Hepatico-jejunostomy was performed using interrupted 4/0 Biosyn sutures while the gastro-jejunostomy or duodenojejunostomy (in PPPD) was performed using continuous 3/0 Polydioxanone (PDS) sutures in 2-layers. One or two surgical drains were placed and the abdomen was closed after ensuring haemostasis.

@@

\subsection{Complications}
The incidence of complications after pancreaticoduodenectomy remains high in spite of a steady decline in postoperative mortality from over 40\% in the 1950's to less than 5\% in most large volume centres around the world.\parencite{deoliveira_assessment_2006, emick_hospital_2006, yeo_six_1997, winter_1423_2006, teh_patient_2009, gouma_rates_2000} Accurate definitions and clinical severity grading are useful in comparing outcomes between centres as well as for research, audit and quality improvement purposes. The consensus definitions of the International Study Group for Pancreatic Surgery (ISGPS) and the Clavien-Dindo classification systems were used during the course of the studies reported here.

@@

Postoperative pancreatic fistula is one of the most dreaded complications after a pancreaticoduodenectomy and can be associated with significant short-term morbidity as well as long-term disability. The reported incidence of postoperative pancreatic fistula varies from 2\% to 30\% after pancreaticoduodenectomy. \parencite{yeo_six_1997, deoliveira_assessment_2006, bassi_postoperative_2005, winter_biochemical_2007, pratt_risk_2008} The variation in reported incidence has been largely due to lack of clear definition of what constituted a postoperative pancreatic fistula. It can be the result of breakdown or poor healing at the pancreatico-jejunostomy/ pancreatico-gastrostomy or may be the result of direct parenchymal leak unrelated to the anastomosis.

It is now generally accepted that 1 in 4 patients will develop a pancreatic fistula as defined by the International Study Group for Pancreatic Fistula (ISGPF) which has published a consensus statement on the definition and grading of postoperative pancreatic fistula.\parencite{bassi_postoperative_2005} A postoperative pancreatic fistula is defined as drain output of any measurable quantity after the third postoperative day with amylase content greater than three times the upper limit of the normal serum amylase value at the laboratory used for testing.

@@

Haemorrhage is often secondary to non-healing of the pancreatico-jejunal anastomosis leading to leakage of amylase-rich pancreatic juices into the retroperitoneum or secondary to intra-abdominal sepsis or bile leak.\parencite{tien_risk_2005, koukoutsis_haemorrhage_2006, choi_delayed_2004, balladur_bleeding_1996} This can then lead to erosion of ligated blood vessels, most commonly the stump of the gastro-duodenal artery. Post-pancreatectomy haemorrhage is usually managed with angiographic embolisation of the bleeding vessel and surgical intervention is only rarely required. The grading of severity of post-pancreatectomy haemorrhage as described by the International Study Group of Pancreatic Surgery\parencite{wente_postpancreatectomy_2007} is shown in Table \ref{table:isgps_pph} on p\pageref{table:isgps_pph}.

\subsubsection{Clavien-Dindo classification of complications}
\input{Tables/c1_clavien_dindo}
A number of other adverse events may occur following pancreaticoduodenectomy including cardiopulmonary complications such as myocardial infarction, cardiac arrhythmias, pneumonia and pleural effusions, wound complications such as wound sepsis and dehiscence, intra-abdominal problems including intra-abdominal sepsis, leakage from the hepatico-jejunostomy or the gastro-jejunostomy, infected fluid collections, renal dysfunction, deep vein thrombosis, pulmonary embolism, etc. The Clavien-Dindo system grades the severity of complications based on the impact of the complication on the management of the patient and has been validated in large numbers of surgical patients.\parencite{clavien_clavien-dindo_2009, dindo_classification_2004} The classification is summarised in Table \ref{table:clavien-dindo} on p\pageref{table:clavien-dindo} and was used to grade complications in this thesis. Complications of grade 3 or greater are considered clinically significant as they require further procedures either under local or general anaesthesia, single or multiple organ support or result in death while grades 1 and 2 are managed with ward-based care including antibiotics, wound management, supplemental enteral or parenteral nutrition and similar interventions.

@@

While surgery is the primary modality of curative treatment in patients with resectable pancreatic cancer, there is clear evidence that adjuvant chemotherapy improves survival and is now the standard of care for patients with pancreatic ductal adenocarcinoma. The European Study Group for Pancreatic Cancer (ESPAC) has conducted some of the largest randomised controlled trials of adjuvant chemotherapy and chemoradiotherapy versus observation after potentially curative surgery and have reported that chemotherapy prolonged survival after surgery while chemoradiotherapy had a harmful effect. \parencite{neoptolemos_adjuvant_2001, neoptolemos_randomized_2004, neoptolemos_adjuvant_2009, neoptolemos_adjuvant_2010}

The ESPAC-1 trial compared the use of chemotherapy with Fluorouracil and Folinic acid or chemoradiotherapy with 20Gy radiotherapy and 5-Fluorouracil (5-FU) versus observation only in patients who underwent potentially curative surgery for pancreatic ductal adenocarcinoma. \parencite{neoptolemos_randomized_2004} The 5-year survival rate in patients who received chemotherapy was 20\% while it was 8\% in patients who did not receive chemotherapy, demonstrating a clear survival benefit for adjuvant chemotherapy. However, adjuvant chemoradiotherapy resulted in 5-year survival dropping from 20\% to 10\% and has largely been abandoned in the adjuvant setting. The ESPAC-3 trial compared the efficacy of 5-FU and Folinic acid with Gemcitabine and reported that Gemcitabine was not superior to 5-FU/Folinic acid. \parencite{neoptolemos_adjuvant_2010}

However, 25\% to 50\% of patients who undergo surgery for pancreatic ductal adenocarcinoma either do not receive or do not complete a full course of adjuvant chemotherapy. \parencite{sohn_resected_2000, neoptolemos_randomized_2004, oettle_adjuvant_2007} A number of factors have been identified for the non-receipt of adjuvant therapy including comorbidity, postoperative complications and favourable pathological features.\parencite{spitz_preoperative_1997, aloia_delayed_2007, oettle_adjuvant_2007, merchant_adjuvant_2009, russ_impact_2009}

@@

The presence of comorbid conditions is associated with increased incidence of adverse outcomes in patients undergoing treatment for pancreatic cancer\parencite{mann_review_2010}. It can also limit the therapeutic options available due to the associated complications or side effects of surgery or adjuvant treatment. \parencite{sandroussi_sociodemographics_2010} Patients with multiple comorbidities are more likely to have higher readmission rates, morbidity and mortality following discharge after pancreaticoduodenectomy. \parencite{schneider_patient_2012}

DeOliveira and co-workers reported that cardiovascular disease was a risk factor not only for overall morbidity but also complication severity after pancreaticoduodenectomy.\parencite{deoliveira_assessment_2006} Cancer cachexia is associated with increased incidence of complications and mortality after pancreaticoduodenectomy \parencite{pausch_cachexia_2012} while obesity is known to be associated with greater incidence and severity of postoperative complications.\parencite{benns_impact_2009} However, existing methods of measuring the impact of comorbidity on physiological fitness are limited and do not adequately predict outcomes after major pancreatic surgery. \parencite{shah_limitations_2012}

@@

The anaerobic threshold (variously described as the lactate threshold or ventilatory threshold) is the point during exercise when oxygen demand by exercising skeletal muscle outstrips supply. Therefore, muscle tissues use anaerobic respiration to supplement aerobic respiration to continue generation of ATP. The resulting metabolic lactic acidosis is almost immediately compensated by the bicarbonate buffer as below: 
\begin{equation} \label{eq:bicarb_buffer}
	H^+ + HCO3^- \Longleftrightarrow H_2CO_3 \Longleftrightarrow H_2O + CO_2
\end{equation}
The resulting excess $CO_2$ is exhaled and is one of the many parameters measured during cardiopulmonary exercise testing. This transition from aerobic to anaerobic respiration may be determined using the V-slope method\parencite{sue_metabolic_1988} or the ventilatory equivalents method.\parencite{beaver_new_1986} Most centres, like ours, use both methods supplemented by information from a variety of other parameters to enable accurate determination of the anaerobic threshold as recommended by the American Thoracic Society/American College of Chest Physicians Statement on cardiopulmonary exercise testing.\parencite{society_ats/accp_2003}

@@

\subsubsection{Ventilatory equivalents method}
The ventilatory equivalents method of determining $\dot{V}_{O_2}$AT involves plotting $\dot{V}_E/\dot{V}_{O_2}$
and $\dot{V}_E/\dot{V}_{CO_2}$ against time, exercise load or $\dot{V}_{O_2}$. The point at which $\dot{V}_E/\dot{V}_{O_2}$ increases while $\dot{V}_E/\dot{V}_{CO_2}$ stays the same or decreases is the anaerobic threshold and this can be seen in Fig. \ref{fig:cpet_9panel}F. Most respiratory physiologists including at our centre use both methods as well as other trends during exercise to determine the anaerobic threshold as well as its validity.

@@

Several studies over the past 2 decades have established the value of cardiopulmonary exercise testing in patients, especially elderly, undergoing major abdominal surgery. Older and co-workers conducted a prospective study of 187 patients over the age of 60 undergoing major abdominal surgery. All patients underwent a symptom-limited exercise test on a cycle ergometer with real-time 12-lead ECG monitoring. The average $\dot{V}_{O_2}$ across all patients was 12.4 ml/min/kg. There were a total of 11 deaths related to cardiovascular causes (5.9\%) and 3 deaths due to non-cardiovascular causes. They reported that 10 out of the 11 deaths due to cardiovascular causes occurred in patients with $\dot{V}_{O_2}AT<11$ml/kg/min (n=55) while only one patient in the group with $\dot{V}_{O_2}AT\geq11$ml/kg/min (n=132) died of cardiovascular complications (18\% vs 0.8\%, $p<0.001$). Eight out of the 10 patients in the low $\dot{V}_{O_2}$AT group who died also had evidence of myocardial ischaemia ($p<0.01$).\parencite{older_preoperative_1993}

In a later, larger prospective study of 548 patients who underwent major abdominal surgery including colorectal and abdominal aneurysm surgery, they used a risk stratification system that combined $\dot{V}_{O_2}AT<11$ml/kg/min, $\dot{V}_E/\dot{V}_{CO_2}>35$ and evidence of myocardial ischaemia during exercise. High risk patients ($\dot{V}_{O_2}AT<11$ml/kg/min, n=153) were admitted to intensive care after surgery and had a mortality due to cardiovascular causes of 4.6\%. Moderate risk group ($\dot{V}_{O_2}AT>11$ml/kg/min and either $\dot{V}_E/\dot{V}_{CO_2}>35$ or evidence of myocardial ischaemia, n=115) was managed in the high dependency unit after surgery and had a cardiovascular mortality of 1.7\%. Low risk patients (with none of the risk factors mentioned above, n=280) were managed on the general surgical ward with no cardiovascular mortality.\parencite{older_cardiopulmonary_1999} Subsequent literature reviews by Older and co-workers have emphasised the value of cardiopulmonary exercise testing in risk assessment as well as optimising perioperative care in the high-risk surgical patient. \parencite{older_preoperative_2000, older_clinical_2004, older_preoperative_2005}

These findings have since been replicated in several other studies although the threshold value of $\dot{V}_{O_2}$AT as well as the inclusion of other CPET parameters to attribute risk differ in these studies. Hightower and co-workers studied 32 patients over the age of 18 undergoing a variety of major elective abdominal surgery. In this heterogeneous cohort, they reported that the percent predicted anaerobic threshold achieved $<$ 75\%, heart rate at the anaerobic threshold and the heart rate response from rest to anaerobic threshold were all independently associated with postoperative morbidity. While it is difficult to extrapolate results from this small heterogeneous cohort to the general surgical population, it would appear that other cardiopulmonary exercise test parameters may have a role in predicting risk in these patients.\parencite{hightower_pilot_2010}

>>>>

Pancreatic ductal adenocarcinoma is the most common cancer of the pancreas. However, the head of the pancreas is anatomically related to several other epithelium lined structures that can also give rise to cancers. These include the distal common bile duct (cholangiocarcinoma), the duodenum (duodenal adenocarcinoma) and the ampulla (ampullary adenocarcinoma). Neuroendocrine tumours (NET) are a collection of a variety of neoplasia that originate from the endocrine portion of the pancreas. Other neoplasia such as intra-ductal papillary neoplasms (IPMN) as well as rare stromal tumours may also occur in this region. Occasionally, chronic pancreatitis may present with features similar to pancreatic cancer and can be morphologically, radiologically and histologically difficult to differentiate from cancer.

Pancreatic cancer is the tenth most common cancer in the UK but the fifth most common cause of cancer death. Only 21\% survive beyond the first year and 5-year survival is 3\% \parencite{cancerresearchuk_cancer_2014}. The majority of patients (80-85\%) with pancreatic cancer present with inoperable disease \parencite{cancerresearchuk_cancer_2014,sener_pancreatic_1999}.

@@

Initial assessment includes a thorough history, clinical examination and blood tests including full blood count, renal function tests, liver function tests and tumour markers, especially CA 19-9. However, definitive diagnosis requires cross-sectional imaging in the form of a contrast-enhanced computerised tomogram (CECT) of the abdomen using a pancreas-specific protocol (a modified form of the portal-venous phase). CECT of the pancreas when combined with CT Thorax also provides accurate information on staging of the disease with regards to metastasis and this can be supplemented by further imaging such as Positron Emission Tomography (PET-CT) or contrast-enhanced Magnetic Resonance Imaging (MRI) of the liver in specific cases. CECT-pancreas is also useful for assessing local resectability with accurate information regarding vascular involvement. Endoscopic Ultrasound (EUS) is used in selected patients to assess vascular involvement and for obtaining tissue samples for histological examination. In jaundiced patients, endoscopic retrograde cholangio pancreatography (ERCP) plays an important role in the alleviation of jaundice by placing stents across the obstructed bile duct, obtaining accurate visualisation of the biliary anatomy as well as obtaining brushings from within the bile duct for cytological examination. The role of preoperative biliary drainage is discussed in more detail in Section \ref{sec:preoperative_biliary_drainage}.

\subsection{Treatment of pancreatic cancer}
Pancreaticoduodenectomy followed by adjuvant chemotherapy offers the only chance of cure in patients with resectable pancreatic cancer who are fit enough to undergo surgery. In patients with unresectable disease or who are not fit to undergo surgery, palliative chemotherapy plays a limited role in prolonging survival. Assessment of resectability of lesions of the head of the pancreas is discussed in the next section while the assessment of patient fitness and the impact of comorbidity are discussed in detail in section \ref{sec:comorbidity_risk_stratification} on p\pageref{sec:comorbidity_risk_stratification}.

\section{Surgical treatment of pancreatic cancer}
Pancreaticoduodenectomy remains a technically challenging and complex surgical procedure.% over a hundred years after its description. 
Excision of cancer of the head of the pancreas was first described as a two-stage operation by a German surgeon, Walther Kausch in 1909 at Augusta-Viktoria-Krankenhaus in Berlin-Schöneberg \parencite{kausch_carcinom_1912}. The operation was further popularised initially as a two-stage procedure by Whipple \parencite{whipple_treatment_1935} before evolving into the current single stage operation by the 1950s \parencite{whipple_rationale_1941, whipple_radical_1950}.

\subsection{Resectability criteria}
\label{sec:resectability_criteria}
Resectable pancreatic cancer is defined as a tumour that does not involve the coeliac axis or the superior mesenteric artery and is not associated with distant metastatic disease. Tumours involving the portal vein or superior mesenteric vein are considered borderline resectable and can still be resected completely (R0) with en-bloc venous resection in selected cases \parencite{fuhrman_rationale_1996}. The role of neoadjuvant therapy such as chemotherapy with or without local radiotherapy and newer treatment modalities such as electroporation \parencite{bower_irreversible_2011} and intra-peritoneal chemotherapy is the subject of ongoing clinical research in patients with locally advanced and/or metastatic disease. Early reports suggest that some of these patients can undergo potentially curative surgery after neo-adjuvant treatment.

@@

A roof-top incision was made to enter the abdominal cavity. After assessing the peritoneal cavity and liver for absence of metastatic disease, an early assessment was made for local resectability. This involved complete Kocherisation of the duodenum to assess the retroperitoneum. Both the superior mesenteric artery and coeliac axis were assessed early for tumour involvement (‘artery-first’ approach). The gastrocolic omentum was divided to enter the lesser sac. The superior mesenteric vein was identified and a retro-pancreatic tunnel was created between the pancreatic neck and the portal vein. If less than half the circumference of the superior mesenteric vein or portal vein was involved, an en-bloc resection was performed with vein repair at the same time. The hepatoduodenal ligament was dissected after a fundus-first cholecystectomy to isolate the common bile duct which was transected after ascertaining the hepatic artery anatomy. The gastro-duodenal artery was divided leaving a short segment on the common hepatic artery should the patient require embolisation of this vessel for post-pancreatectomy haemorrhage. The stomach (classical Whipple) or the first part of the duodenum (pylorus-preserving pancreaticoduodenectomy) were divided using a stapling device. The pancreatic neck was transected using Harmonic scalpel, diathermy or cold blade depending on surgeon preference. The jejunum was divided at the first jejunal arcade and the jejunal mesentery was divided using the Harmonic ultrasonic scalpel. The retroportal lamina was divided preserving the superior mesenteric artery.

Reconstruction was performed as follows: either a pancreatico-jejunostomy was fashioned using 4-0 Biosyn sutures in a two-layer duct-to-mucosa technique or a pancreatico-gastrostomy was created using 3/0 Biosyn sutures placed in a similar manner. Hepatico-jejunostomy was performed using interrupted 4/0 Biosyn sutures while the gastro-jejunostomy or duodenojejunostomy (in PPPD) was performed using continuous 3/0 Polydioxanone (PDS) sutures in 2-layers. One or two surgical drains were placed and the abdomen was closed after ensuring haemostasis.

@@

\subsection{Complications}
The incidence of complications after pancreaticoduodenectomy remains high in spite of a steady decline in postoperative mortality from over 40\% in the 1950's to less than 5\% in most large volume centres around the world \parencite{deoliveira_assessment_2006, emick_hospital_2006, yeo_six_1997, winter_1423_2006, teh_patient_2009, gouma_rates_2000}. Accurate definitions and clinical severity grading are useful in comparing outcomes between centres as well as for research, audit and quality improvement purposes. The consensus definitions of the International Study Group for Pancreatic Surgery (ISGPS) and the Clavien-Dindo classification systems were used during the course of the studies reported here.

@@

Postoperative pancreatic fistula is one of the most dreaded complications after a pancreaticoduodenectomy and can be associated with significant short-term morbidity as well as long-term disability. The reported incidence of postoperative pancreatic fistula varies from 2\% to 30\% after pancreaticoduodenectomy \parencite{yeo_six_1997, deoliveira_assessment_2006, bassi_postoperative_2005, winter_biochemical_2007, pratt_risk_2008}. The variation in reported incidence has been largely due to lack of clear definition of what constituted a postoperative pancreatic fistula. It can be the result of breakdown or poor healing at the pancreatico-jejunostomy/ pancreatico-gastrostomy or may be the result of direct parenchymal leak unrelated to the anastomosis.

It is now generally accepted that 1 in 4 patients will develop a pancreatic fistula as defined by the International Study Group for Pancreatic Fistula (ISGPF) which has published a consensus statement on the definition and grading of postoperative pancreatic fistula \parencite{bassi_postoperative_2005}. A postoperative pancreatic fistula is defined as drain output of any measurable quantity after the third postoperative day with amylase content greater than three times the upper limit of the normal serum amylase value at the laboratory used for testing.

@@

Haemorrhage is often secondary to non-healing of the pancreatico-jejunal anastomosis leading to leakage of amylase-rich pancreatic juices into the retroperitoneum or secondary to intra-abdominal sepsis or bile leak \parencite{tien_risk_2005, koukoutsis_haemorrhage_2006, choi_delayed_2004, balladur_bleeding_1996}. This can then lead to erosion of ligated blood vessels, most commonly the stump of the gastro-duodenal artery. Post-pancreatectomy haemorrhage is usually managed with angiographic embolisation of the bleeding vessel and surgical intervention is only rarely required. The grading of severity of post-pancreatectomy haemorrhage as described by the International Study Group of Pancreatic Surgery \parencite{wente_postpancreatectomy_2007} is shown in Table \ref{table:isgps_pph} on p\pageref{table:isgps_pph}. .

\subsubsection{Clavien-Dindo classification of complications}
\input{Tables/c1_clavien_dindo}
A number of other adverse events may occur following pancreaticoduodenectomy including cardiopulmonary complications such as myocardial infarction, cardiac arrhythmias, pneumonia and pleural effusions, wound complications such as wound sepsis and dehiscence, intra-abdominal problems including intra-abdominal sepsis, leakage from the hepatico-jejunostomy or the gastro-jejunostomy, infected fluid collections, renal dysfunction, deep vein thrombosis, pulmonary embolism, etc. The Clavien-Dindo system grades the severity of complications based on the impact of the complication on the management of the patient and has been validated in large numbers of surgical patients \parencite{clavien_clavien-dindo_2009, dindo_classification_2004}. The classification is summarised in Table \ref{table:clavien-dindo} on p\pageref{table:clavien-dindo}. and was used to grade complications in this thesis. Complications of grade 3 or greater are considered clinically significant as they require further procedures either under local or general anaesthesia, single or multiple organ support or result in death while grades 1 and 2 are managed with ward-based care including antibiotics, wound management, supplemental enteral or parenteral nutrition and similar interventions.

@@

While surgery is the primary modality of curative treatment in patients with resectable pancreatic cancer, there is clear evidence that adjuvant chemotherapy improves survival and is now the standard of care for patients with pancreatic ductal adenocarcinoma. The European Study Group for Pancreatic Cancer (ESPAC) has conducted some of the largest randomised controlled trials of adjuvant chemotherapy and chemoradiotherapy versus observation after potentially curative surgery and have reported that chemotherapy prolonged survival after surgery while chemoradiotherapy had a harmful effect \parencite{neoptolemos_adjuvant_2001, neoptolemos_randomized_2004, neoptolemos_adjuvant_2009, neoptolemos_adjuvant_2010}.

The ESPAC-1 trial compared the use of chemotherapy with Fluorouracil and Folinic acid or chemoradiotherapy with 20Gy radiotherapy and 5-Fluorouracil (5-FU) versus observation only in patients who underwent potentially curative surgery for pancreatic ductal adenocarcinoma \parencite{neoptolemos_randomized_2004}. The 5-year survival rate in patients who received chemotherapy was 20\% while it was 8\% in patients who did not receive chemotherapy, demonstrating a clear survival benefit for adjuvant chemotherapy. However, adjuvant chemoradiotherapy resulted in 5-year survival dropping from 20\% to 10\% and has largely been abandoned in the adjuvant setting. The ESPAC-3 trial compared the efficacy of 5-FU and Folinic acid with Gemcitabine and reported that Gemcitabine was not superior to 5-FU/Folinic acid. \parencite{neoptolemos_adjuvant_2010}

However, 25\% to 50\% of patients who undergo surgery for pancreatic ductal adenocarcinoma either do not receive or do not complete a full course of adjuvant chemotherapy \parencite{sohn_resected_2000, neoptolemos_randomized_2004, oettle_adjuvant_2007}. A number of factors have been identified for the non-receipt of adjuvant therapy including comorbidity, postoperative complications and favourable pathological features \parencite{spitz_preoperative_1997, aloia_delayed_2007, oettle_adjuvant_2007, merchant_adjuvant_2009, russ_impact_2009}.

@@

The presence of comorbid conditions is associated with increased incidence of adverse outcomes in patients undergoing treatment for pancreatic cancer \parencite{mann_review_2010}. It can also limit the therapeutic options available due to the associated complications or side effects of surgery or adjuvant treatment \parencite{sandroussi_sociodemographics_2010}. Patients with multiple comorbidities are more likely to have higher readmission rates, morbidity and mortality following discharge after pancreaticoduodenectomy \parencite{teh_patient_2009}.

DeOliveira and co-workers reported that cardiovascular disease was a risk factor not only for overall morbidity but also complication severity after pancreaticoduodenectomy \parencite{deoliveira_assessment_2006}. Obesity is known to be associated with greater incidence and severity of postoperative complications \parencite{benns_impact_2009}. However, existing methods of measuring the impact of comorbidity on physiological fitness are limited and do not adequately predict outcomes after major pancreatic surgery \parencite{de_castro_evaluation_2009}.

@@

The anaerobic threshold (variously described as the lactate threshold or ventilatory threshold) is the point during exercise when oxygen demand by exercising skeletal muscle outstrips supply. Therefore, muscle tissues use anaerobic respiration to supplement aerobic respiration to continue generation of ATP. The resulting metabolic lactic acidosis is almost immediately compensated by the bicarbonate buffer as below: 
\begin{equation} \label{eq:bicarb_buffer}
	H^+ + HCO3^- \Longleftrightarrow H_2CO_3 \Longleftrightarrow H_2O + CO_2
\end{equation}
The resulting excess $CO_2$ is exhaled and is one of the many parameters measured during cardiopulmonary exercise testing. This transition from aerobic to anaerobic respiration may be determined using the V-slope method \parencite{sue_metabolic_1988} or the ventilatory equivalents method \parencite{beaver_new_1986}. Most centres, like ours, use both methods supplemented by information from a variety of other parameters to enable accurate determination of the anaerobic threshold as recommended by the American Thoracic Society/American College of Chest Physicians Statement on cardiopulmonary exercise testing \parencite{society_ats/accp_2003}.

@@

\subsubsection{Ventilatory equivalents method}
The ventilatory equivalents method of determining $\dot{V}_{O_2}$AT involves plotting $\dot{V}_E/\dot{V}_{O_2}$
and $\dot{V}_E/\dot{V}_{CO_2}$ against time, exercise load or $\dot{V}_{O_2}$. The point at which $\dot{V}_E/\dot{V}_{O_2}$ increases while $\dot{V}_E/\dot{V}_{CO_2}$ stays the same or decreases is the anaerobic threshold and this can be seen in Fig. \ref{fig:cpet_9panel}F. Most respiratory physiologists including those at our centre use both methods as well as other trends during exercise to determine the anaerobic threshold as well as its validity.

@@

Several studies over the past 2 decades have established the value of cardiopulmonary exercise testing in patients, especially elderly, undergoing major abdominal surgery. Older and co-workers conducted a prospective study of 187 patients over the age of 60 undergoing major abdominal surgery. All patients underwent a symptom-limited exercise test on a cycle ergometer with real-time 12-lead ECG monitoring. The average $\dot{V}_{O_2}$ across all patients was 12.4 ml/min/kg. There were a total of 11 deaths related to cardiovascular causes (5.9\%) and 3 deaths due to non-cardiovascular causes. They reported that 10 out of the 11 deaths due to cardiovascular causes occurred in patients with $\dot{V}_{O_2}AT<11$ml/kg/min (n=55) while only one patient in the group with $\dot{V}_{O_2}AT\geq11$ml/kg/min (n=132) died of cardiovascular complications (18\% vs 0.8\%, $p<0.001$). Eight out of the 10 patients in the low $\dot{V}_{O_2}$AT group who died also had evidence of myocardial ischaemia ($p<0.01$) \parencite{older_preoperative_1993}.

In a later, larger prospective study of 548 patients who underwent major abdominal surgery including colorectal and abdominal aneurysm surgery, they used a risk stratification system that combined $\dot{V}_{O_2}AT<11$ml/kg/min, $\dot{V}_E/\dot{V}_{CO_2}>35$ and evidence of myocardial ischaemia during exercise. High risk patients ($\dot{V}_{O_2}AT<11$ml/kg/min, n=153) were admitted to intensive care after surgery and had a mortality due to cardiovascular causes of 4.6\%. Moderate risk group ($\dot{V}_{O_2}AT>11$ml/kg/min and either $\dot{V}_E/\dot{V}_{CO_2}>35$ or evidence of myocardial ischaemia, n=115) was managed in the high dependency unit after surgery and had a cardiovascular mortality of 1.7\%. Low risk patients (with none of the risk factors mentioned above, n=280) were managed on the general surgical ward with no cardiovascular mortality \parencite{older_cardiopulmonary_1999}. Subsequent literature reviews by Older and co-workers have emphasised the value of cardiopulmonary exercise testing in risk assessment as well as optimising perioperative care in the high-risk surgical patient. \parencite{older_preoperative_2000, older_clinical_2004, older_preoperative_2005}.

These findings have since been replicated in several other studies although the threshold value of $\dot{V}_{O_2}$AT as well as the inclusion of other CPET parameters to attribute risk differ in these studies. Hightower and co-workers studied 32 patients over the age of 18 undergoing a variety of major elective abdominal surgery. In this heterogeneous cohort, they reported that the percent predicted anaerobic threshold achieved $<$ 75\%, heart rate at the anaerobic threshold and the heart rate response from rest to anaerobic threshold were all independently associated with postoperative morbidity. While it is difficult to extrapolate results from this small heterogeneous cohort to the general surgical population, it would appear that other cardiopulmonary exercise test parameters may have a role in predicting risk in these patients \parencite{hightower_pilot_2010}.

Snowden and co-workers reported in a study of 171 patients of who 123 underwent surgery, that $\dot{V}_{O_2}AT<10.1$ ml/kg/min was associated with not only cardiovascular complications but also with other complications including pulmonary, renal, gastrointestinal, infective and haematological complications. One of the strengths of this study was the fact that the clinicians treating these patients were blinded to the results of preoperative cardiopulmonary exercise testing results thus avoiding management bias. They also included other cardiopulmonary exercise test parameters such as $\dot{V}_{O_2}$Peak and $\dot{V}_E/\dot{V}_{CO_2}$ at the anaerobic threshold in addition to $\dot{V}_{O_2}$AT in their analysis and showed that $\dot{V}_{O_2}$AT was more predictive of postoperative morbidity than POSSUM derived morbidity, cardiac risk scoring index or a validated activity questionnaire. However, this study also included a heterogeneous cohort of diseases including vascular, pancreatic and hepatobiliary disorders and sarcomas \parencite{snowden_submaximal_2010}.
